Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-12-06 code/maps.qmd

```{r}
# For printing on screen: xmin = 244000, ymax = 4242500, xmax = 309000, ymin = 4142000
bbox <- st_bbox(c(xmin = 245000, ymax = 4231000, xmax = 308000, ymin = 4151000), 
                crs = st_crs(yose_boundary)) %>%  
  st_as_sfc()
hillshade_yose_crop <- st_crop(hillshade_90m, bbox)
hillshade_yose_crop <- st_transform(hillshade_yose_crop, 26911)
```

@@

```{r}
translocation_points_map <- tm_shape(hillshade_yose_crop) + 
  tm_rgb(r = 1, g = 2, b = 3) +
  tm_shape(yose_boundary) + 
  tm_borders(col = "black", lwd = 2) +
  tm_shape(translocation_points) + 
  tm_symbols(size = 0.9, 
             shape = "donor_id",
             col = "type", 
             legend.shape.show = FALSE, 
             palette = c("#0000FF", "#C0FF3E"), # tmap only recognizes basic color names (e.g., "blue" but not "lightblue") - use HEX instead
             title.col = "Population type") +
  tm_layout(legend.position = c("right", "top"),
            legend.bg.color = "white",
            legend.title.size = 1.5,
            legend.text.size = 1.2) +
  tm_compass(type = "4star",
             position = c(0.80, 0.11),
             size = 5) +
  tm_scale_bar(breaks = c(0, 10, 20), 
               text.size = 1.2, 
               position = c("right", "bottom"))

>>>>

```{r}
# For printing on screen: xmin = 244000, ymax = 4242500, xmax = 309000, ymin = 4142000
bbox <- st_bbox(c(xmin = 245000, ymax = 4231000, xmax = 311000, ymin = 4151000), 
                crs = st_crs(yose_boundary)) %>%  
  st_as_sfc()
hillshade_yose_crop <- st_crop(hillshade_90m, bbox)
hillshade_yose_crop <- st_transform(hillshade_yose_crop, 26911)
```

@@

```{r}
#translocation_points <- translocation_points %>% 
# mutate(
#    label = case_when(
#      site_id == 70567 ~ 4, 
#      site_id == 72996 ~ 1
#    ))

translocation_points_map <- tm_shape(hillshade_yose_crop) + 
  tm_rgb(r = 1, g = 2, b = 3) +
  tm_shape(yose_boundary) + 
  tm_borders(col = "#737373", lwd = 2) +
  tm_shape(translocation_points) + 
  tm_symbols(size = 4, 
             shape = "donor_id",
             col = "type", 
             legend.shape.show = FALSE, 
             palette = c("#87CEFA", "#C0FF3E"), # tmap only recognizes basic color names (e.g., "blue" but not "lightblue") - use HEX instead
             title.col = "Population type") +
  tm_layout(legend.position = c("right", "top"),
            legend.bg.color = "white",
            legend.title.size = 1.5,
            legend.text.size = 1.2) +
  tm_text(text = "site_id",
          auto.placement = TRUE,
          size = 1.1) +
  tm_compass(type = "4star",
             position = c(0.80, 0.11),
             size = 5) +
  tm_scale_bar(breaks = c(0, 10, 20), 
               text.size = 1.2, 
               position = c("right", "bottom"))

### 2023-12-06 code/translocation_survival_analysis_supp.Rmd

```{r}
d1 %>% 
  mutate(period = factor(period, levels = c("before", "after"))) %>% 
  ggplot(aes(x = period, y = log10(bd_load + 1))) +
          geom_boxplot()

d1 %>% 
  mutate(period = factor(period, levels = c("before", "after"))) %>% 
  ggplot(aes(x = period, y = log10(bd_load + 1))) +
    geom_boxplot() + 
    labs(x = "Period (relative to translocation)", y = (expression(Bd~load~(log["10"]~copies+1)))) +
    facet_grid(. ~ site_id1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          strip.text.x = element_text(size = 8),
          panel.grid = element_blank())
ggsave(here::here("doc", "manuscript", "figures", "bdload_beforeafter.png"), width = 6, height = 3)
```

>>>>

```{r}
smplsize <- d1 %>% 
  count(site_id1, period)

d1 %>% 
  mutate(period = factor(period, levels = c("before", "after"))) %>% 
  filter(interval < 2) %>% 
  ggplot(aes(x = period, y = log10(bd_load + 1))) +
    geom_boxplot() + 
    labs(x = "Period (relative to translocation)", y = (expression(Bd~load~(log["10"]~copies+1)))) +
    coord_cartesian(ylim = c(-0.5, NA)) +
    facet_grid(. ~ site_id1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          strip.text.x = element_text(size = 8),
          panel.grid = element_blank()) +
    geom_text(data = smplsize, aes(x = period, y = -0.4, label = n), size = 2.2, vjust = 1)

ggsave(here::here("doc", "manuscript", "figures", "bdload_beforeafter.png"), width = 6, height = 3)
```

### 2023-12-06 doc/manuscript/translocation.qmd

---
bibliography: translocation.bib
csl: pnas.csl
# format: docx
format :
  # latex:
  pdf:
#    cite-method: natbib
    include-in-header:
      text: |
        \usepackage{blkarray}
editor: visual
editor_options:
  markdown:
    wrap: sentence
---

@@

Reintroduction of resistant/tolerant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs.
In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
Results from the population viability model were also encouraging.
In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year 1 juvenile survival probability was greater than 0.10).
In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.

Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2018] and as a result experience high mortality.
This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
Therefore, although adult MYL frogs appear relatively resistant/tolerant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].

@@

For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
The populations that we selected contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles.
These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
Recipient sites to which frogs were translocated were chosen based on previous *R. sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].

We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
Frogs were transported to the recipient site either on foot or via helicopter.
Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
However, when the number of frogs detected on the first survey day was zero or near zero, we typically conducted only a single-day CMR survey.
When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.

@@

We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
The possible states of individuals include "not recruited", "alive", and "dead".
The possible observations of individuals include "detected" and "not detected".
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).

>>>>

---
bibliography: translocation.bib
csl: pnas.csl
format: pdf
# format: docx
format :
  # latex:
  pdf:
#    cite-method: natbib
    include-in-header:
      text: |
        \usepackage{blkarray}
editor: visual
editor_options:
  markdown:
    wrap: sentence
---

<!--# Drop "evolutionary rescue" from title? -->

@@

Reintroduction of resistant/tolerant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs <!--# Add to this to deal with Reviewer 2-Comment 4.-->.
In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
Results from the population viability model were also encouraging.
In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction $<$ 0.5 when year 1 juvenile survival probability was greater than 0.10).
In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
<!--# Mention that 1-year survival can be used as an accurate proxy of long-term persistence.  -->

Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2019] and as a result experience high mortality.
This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
Therefore, although adult MYL frogs appear relatively resistant/tolerant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].

@@

For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected using several years of visual encounter surveys (VES) and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
The populations that we selected contained several hundred *R. sierrae* adults and hundreds or thousands of tadpoles.
These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
As is typical for recovering MYL frog populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected (log~10~(load) $\approx$ 5.78 copies)[@vredenburg2010; @joseph2018].
Recipient sites to which frogs were translocated were chosen based on previous *R. sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
At the beginning of this study, we had a relatively limited understanding of these habitat quality characteristics. 
In subsequent years, we improved our site selection process by incorporating new information about important habitat features (in particular, overwinter habitats such as submerged boulders and overhanging banks). *R. sierrae* were absent from all recipient sites prior to the first translocation.

We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
Frogs were transported to the recipient site either on foot or via helicopter.
Following release, translocated populations were visited approximately once per month during the summer active season (generally July-August; median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
However, when the number of frogs detected on the first survey day was zero or near zero, we typically conducted only a single-day CMR survey.
When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.

@@

We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
The possible states of individuals include "not recruited", "alive", and "dead".
The possible observations of individuals include "detected" and "not detected".
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
We also assume that new recruits were a result of within-site reproduction and not immigration from adjacent populations because no *R. sierrae* populations were present within several kilometers of the translocation sites. 
For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
